---
title: "Introduction to OpenTripPlanner - creating a bespoke multi-modal route planner"
author: "Marcus Young"
date: "11 November 2017"
output:
  html_document:
  highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction


## Data and applications

The datasets and applications used for this project are available from a GitHub repository and can be downloaded in a single zip file from here: [https://github.com/marcusyoung/otp-tutorial/data.zip](https://github.com/marcusyoung/otp-tutorial/data.zip).

Once you have extracted this folder, you should see the following files:

* **otp.jar**, the OpenTripPlanner application, in a single runnable JAR file. This has been compiled from a modified version of the source code with average road speeds and traversal rules that are more appropriate for the UK.

* **fn_otp_api.R**, a set of R functions that can be used to query the OTP API and process the response.

* **router-config.json**, a configuration file to be used with OTP

* **greater-manchester-osm.pbf**, an extract from OpenStreetMap (OSM) covering the Greater Manchester area, obtained from: [http://download.geofabrik.de/europe/great-britain/england/greater-manchester.html](http://download.geofabrik.de/europe/great-britain/england/greater-manchester.html).

* **rail-gtfs.zip**, a GTFS feed for UK national rail services, limited to operators that provide services in the Greater Manchester area. Based on a full feed (dated 26/06/17) obtained from: [http://transitfeeds.com/p/association-of-train-operating-companies/284](http://transitfeeds.com/p/association-of-train-operating-companies/284).


* **tfgm-gtfs.zip**, a GTFS feed for transit services in Greater Manchester (dated 28/04/17), provided by Transport for Greater Manchester (TfGM) and obtained from: [http://transitfeeds.com/p/transport-for-greater-manchester/224](http://transitfeeds.com/p/transport-for-greater-manchester/224).

* **aircoach-gtfs.zip**, a GTFS feed for a fictional coach link service between Wigan and Manchester Airport.

* **gm_lsoa_polygons.\***, a Shapefile containing polygons for the 1,740 Lower Layer Super Output Areas (LSOA) that cover the Greater Manchester area.

* **gm_lsoa_centroids.csv**, a CSV file containing atttributes and centroid coordinates for the Greater Manchester LSOAs. 




## Part 1 - Setting up and testing an OTP instance (30 mins)



### Create an OTP Graph

An OTP graph specifies every location in the region covered and how to travel between them, and is compiled by OTP using OSM data for the street and path network (used for walk, bicycle and drive modes) and GTFS data for transit scheduling. It can also incorporate a digital terrain model (provided in GeoTIFF format) which is 'draped' over the street network and can be used in routing - for example to request a flatter cycling route.

Our first task is to create the folder and file structure expected by OTP. This requires a base directory called 'otp' which contains the OTP JAR file, followed by a subdirectory called 'graphs'. Each subdirectory created under 'graphs' is known as an OTP router, and contains all the files required to build a graph. A single OTP instance can host several routers - for example covering different regions, or containing alternative data for the same region. 

Create a router called 'current', and include the GTFS and OSM files for the road network and current transit schedules for Greater Manchester, along with router-config.json, as shown below:   

```{r, engine = 'bash', eval = FALSE}
/otp
 └── otp.jar
 └── graphs
     └── current
         ├── rail-gtfs.zip
         ├── tfgm-gtfs.zip
         ├── greater-manchester-osm.pbf
         └── router-config.json
         
```

With the structure created and the files in place, we can now build the graph. We do this by running OTP and specifiying the 'build' option. OTP runs within a Java virtual machine (JVM), which is provided by the Java runtime environment (JRE). Before continuing, run `java -version` from the command prompt or terminal to check that you have version 1.8 of the JVM installed. If you do not have the correct JRE installed you will need to install it for your operating system from: [http://www.oracle.com/technetwork/java/javase/downloads/jre8-downloads-2133155.html](http://www.oracle.com/technetwork/java/javase/downloads/jre8-downloads-2133155.html])

Open the command prompt or terminal and change to the 'otp' directory you created earlier. Then run the following command:

```{r, engine = 'bash', eval = FALSE}
java -Xmx2G -jar otp.jar --build graphs/current
```

The java option -Xmx2G allocates up to 2GB of RAM to the JVM. If you do not have this much free RAM available OTP will likely fail during the build process. If you get a message similar to the one below, and think you have enough RAM available, you may need to install the 64-bit version of JRE, available at the link above.

```{r, engine = 'dot', eval = FALSE}
Error occurred during initialization of VM
Unable to allocate ... bitmaps for parallel garbage collection for the requested ... heap.
Error: Could not create the Java Virtual Machine.
```

The build process should take 2-3 minutes to complete. Don't worry about the errors that flash up during the process, these mostly relate to railway station stops that are located outside of the Greater Manchester are, and there is no road network to link them to. Once complete, you will see a file called 'Graph.obj' in the 'current' directory.


### Start up OTP server

To start up your OTP server, using the router called `current' and the graph you create above



If you have time, create an additional router called 'new' and (you can always do this during Part 2 below)


## Task 2 - generating travel time isochrones



## Task 3 - automating OTP API lookups using R